---
title: "Durham NC Bicycle Safety data CodeForAmerica project"
author: "David Hopp "
date: "2016-05-09 "
output: html_document
---  

R markdown script to perform all operations  
leading to the final output csv file rawDataFixedMiss.csv  
 
## Step 0. Make sure required packages loaded
```{r step00, results='hide'}
require(plyr)
require(dplyr)
require(stringr)
require(reshape2)
require(RCurl)
```

# Step 1. Define Functions  

```{r step01}
testFix <- function(expr, doWhat=FALSE, compareto=NA) {
#
# Use testFix to test whether a change was properly applied to some data structure.
# 'expr', must be an expression nominally prepared by expression().
# insist on expr and compareto being of same class, this may require
#	something like as.integer() in the call.
# setting doWhat TRUE requires a value for compareto.
# set doWhat FALSE if no comparison is to be made.
#
# if doWhat TRUE comparison is made and is the result of identical(eval(expr),compareto):
# Returns a list with first item TRUE or FALSE indicating success or failure
# if TRUE second and third items are blanks ""
# if FALSE second item is a word, at present ERROR or WARNING
# and third item is a diagnostic message
#
# if doWhat FALSE no comparison is made (useful for verifying eval(expr))
# Returns a list with first item TRUE
# second item is eval(expr)
# third item is class(eval(expr))
#
	res <- eval(expr)
	eclass <- class(res)
	if (!doWhat) {
		return( list(TRUE, res, eclass) )
	} else {
		cclass <- class(compareto)
		if ( !identical(eclass,cclass) ) {
			return( list(FALSE, "ERROR", paste("res class:",eclass,"compareto class:",cclass)) )
		}
		if ( identical(res,compareto) ) {
			return( list(TRUE,"","") )
		}
		return( list(FALSE, "FAIL", "classes same but values different") )
	}
}
```

## Step 2. Read the csv file acquired from the github project  
 
set "download" TRUE if file to be read from remote host  
set "download" FALSE if skipping remote read and using local csv file  
Presumption is that all files will be in the current working directory  
If you read the remote file it will be placed in strDownload in its original form 
and also a csv representation of it will be written to localCopyOfRemoteFile_csv  
If you read the local copy it must be in localCopyOfRemoteFile_csv

```{r step02, error=FALSE}
# borrowed from Brian:
#
strDownload <- "CrashDataFromRemote.csv.gz"  
localCopyOfRemoteFile_csv <- "RawDataFromRemote.csv"
bikeSafetyURL <- "https://github.com/BikeSafety/BikeSafety/blob/master/data/North_Carolina_Bicycle_Crash_Data.csv.gz?raw=true"
# does URL exist and is accessible?
if ( !RCurl::url.exists( bikeSafetyURL ) ) {
	stop( "URL does not exist or is not accessible." )
}
#
download <- FALSE
#
if (download) {
  download.file( bikeSafetyURL, strDownload, mode="wb" )
  rawData <- read.csv( gzfile(strDownload), stringsAsFactors = FALSE )
  unlink(strDownload)
  # write file as csv
  rawData %>% write.csv( localCopyOfRemoteFile_csv, row.names = FALSE )
} else {
	if ( !exists("localCopyOfRemoteFile_csv") | is.na(localCopyOfRemoteFile_csv) | nchar(str_trim(localCopyOfRemoteFile_csv))==0 ) {
		stop( "Local file name not provided." )
	}
	if ( !file.exists(localCopyOfRemoteFile_csv) ) {
		stop( paste("Local file does not exist:", localCopyOfRemoteFile_csv) )
	}
	# at last
	rawData <- read.csv( localCopyOfRemoteFile_csv, sep=",", stringsAsFactors=FALSE )
}
#
nrow(rawData)
str(rawData)
```

## Step 3. Look at some data values
```{r step03, error=FALSE}
 plyr::count(rawData, vars="BikeAge_Gr")
# notice the erroneous values at lines 3 and 4
```

## Step 4. Look at categorical data and see if any "odd" entries
```{r step04, error=FALSE}
# As a result of observation the rawData categorical variables (columns) - keeping row number OBJECTID - are:
catcol_rawData <- c("OBJECTID", "AmbulanceR","BikeAge_Gr","Bike_Alc_D","Bike_Dir","Bike_Injur","Bike_Pos",
"Bike_Race","Bike_Sex","City","County","CrashAlcoh","CrashDay","Crash_Grp","Crash_Hour",
"Crash_Loc","Crash_Mont","Crash_Type","Crash_Ty_1","Crash_Year","Crsh_Sevri","Developmen","DrvrAge_Gr",
"Drvr_Age","Drvr_Alc_D","Drvr_EstSp","Drvr_Injur","Drvr_Race","Drvr_Sex","Drvr_VehTy","ExcsSpdInd","Hit_Run",
"Light_Cond","Locality","Num_Lanes","Num_Units","Rd_Charact","Rd_Class","Rd_Conditi",
"Rd_Config","Rd_Defects","Rd_Feature","Rd_Surface","Region","Rural_Urba","Speed_Limi","Traff_Cntr","Weather",
"Workzone_I" )
#
# work with only the categorical columns
rawData_cat <- dplyr::select( rawData, one_of(catcol_rawData) )
#
str(rawData_cat)
 # compute the values and counts of the categorical columns
 # these will require visual review to identify any odd entries
 #
 rawData_catmelt <- reshape2::melt(rawData_cat,id="OBJECTID")
 nrow(rawData_catmelt)
 rawData_catcount <- plyr::count( rawData_catmelt, vars=c("variable","value") )
 nrow(rawData_catcount)
```
 
## Step 5. Observations  

  Handle BLANKS leading and trailing  (will use str_trim to remove them)  
 
 There may be some entries that have trailing blanks  
 one way to find those is here:
```{r step05, error=FALSE}
 rawData_catmelt_trblank <- dplyr::filter(rawData_catmelt, str_detect(value," +$"))
 nrow(rawData_catmelt_trblank)
 # which var has trailing blanks?
 unique(rawData_catmelt_trblank$variable)
 #
 # some may start with a blank
 rawData_catmelt_strblank <- dplyr::filter(rawData_catmelt, str_detect(value,"^ +"))
 nrow(rawData_catmelt_strblank)
 #
 reshape2::dcast(rawData_catmelt_strblank,value~variable)
```

# Here are some significant observations:  
There are 4 definitely *undependable* records in rawData: 
				2902, 3188, 4108, 4620  
There are other records that look _unreliable_:  
				371, 846, 942, 1247, 1271, 1777, 2101, 2130, 2789   
   

## Step 6. There are records with *empty* columns or pseudo-codes for missing values:  
(same for trimmed value)
```{r step06, error=FALSE}
rawData_catcount[rawData_catcount$value=="",]
# (same for trimmed value)
rawData_catcount[rawData_catcount$value=="/Missing",]
# (same for trimmed value)
rawData_catcount[rawData_catcount$value=="Missing",]
```
Note that Crash_Loc has `r rawData_catcount[rawData_catcount$variable=="Crash_Loc" & stringr::str_trim(rawData_catcount$value)=="Location","freq"]` entries with "Location"  

## Step 7. "Fix" rawData  
result will be df **rawDataFixedMiss**  

 SPECIAL TREATMENT OF COLUMNS Longitude and Location  
 zeroes appear in both, as 0 and 0.00000 for Longitude  
 and (0, 0) for Location (watch out for the space)  
 these are best replaced with NA but that depends on   
 how Location is handled in the subsequent processing programs.  
 For now, replace with NA  
  
```{r step07, error=FALSE}
rawDataFixed <- rawData
t1 <- nrow(rawDataFixed)
# remove suspicious rows 
rowsToRemove <- c(2902, 3188, 4108, 4620,371, 846, 942, 1247, 1271, 1777, 2101, 2130, 2789)
t1 <- t1-length(rowsToRemove)
rawDataFixed <- rawDataFixed[ -rowsToRemove, ,drop=FALSE]
testFixReturn <- testFix(expression(nrow(rawDataFixed)), TRUE, as.integer(t1))
if (testFixReturn[[1]]) {
	print("correct number of rows deleted")
} else {
	stop( paste(testFixReturn[[2]],testFixReturn[[3]]) )
}
#
# TRIM TEXT FIELDS
mycols <- c("Bike_Injur","Crash_Loc","Crash_Type","Crsh_Sevri","Drvr_Injur","Drvr_VehTy")
rawDataFixed[mycols] <- lapply( rawDataFixed[mycols], function(x) stringr::str_trim(x) )
nrow(rawDataFixed)
```

## Step 8. Go through prep steps for categorical variables  
using rawDataFixed as the basis  
```{r step08, error=FALSE}
rawDataFixed_cat <- dplyr::select( rawDataFixed, one_of(catcol_rawData) )
#
rawDataFixed_catmelt <- reshape2::melt( rawDataFixed_cat,id="OBJECTID" )
rawDataFixed_catcount <- plyr::count( rawDataFixed_catmelt, vars=c("variable","value") )
#
nrow(rawDataFixed_catmelt)
nrow(rawDataFixed_catcount)
rawDataFixed_catmelt_trblank <- dplyr::filter( rawDataFixed_catmelt, str_detect(value," +$") )
#
# the result should be zero rows:
testFixReturn <- testFix( expression( nrow( rawDataFixed_catmelt_trblank ) ), TRUE, as.integer(0) )
if (testFixReturn[[1]]) {
	print("correct: zero rows remaining to be trimmed")
} else {
	stop( paste(testFixReturn[[2]],testFixReturn[[3]]) )
}
```

## Step 9. Fix missing values  
replace with NA  
```{r step09, error=FALSE}
 rawDataFixedMiss <- rawDataFixed
 rawDataFixedMiss[rawDataFixedMiss=="" | rawDataFixedMiss=="." | rawDataFixedMiss=="Missing" | rawDataFixedMiss=="/Missing"] <- NA
 nrow(rawDataFixedMiss)
```

## Step 10. Fix Longitude and Location - replace zeros with NA   
```{r step10, error=FALSE}
 sum(rawDataFixedMiss$Location=="(0, 0)")

 rawDataFixedMiss$Location[rawDataFixedMiss$Location=="(0, 0)"] <- NA
 sum(is.na(rawDataFixedMiss$Location))
 sum(rawDataFixedMiss$Longitude==0)
 rawDataFixedMiss$Longitude[rawDataFixedMiss$Longitude==0] <- NA
 sum(is.na(rawDataFixedMiss$Longitude))
```

## Step 11. Fix BikeAge_Gr when it contains "10-Jun" or "15-Nov"
Note that BikeAge_Gr has
 `r rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value=="10-Jun","freq"]` instances of "10-Jun" and 
 `r rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value=="15-Nov","freq"]` instances of "15-Nov" 
```{r step11, error=FALSE}
temp <- rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value %in% c("10-Jun","15-Nov"),"freq"]
if ( length(temp) > 0 ) {
	rawDataFixedMiss <- rawDataFixedMiss %>% dplyr::mutate( BikeAge_Gr=ifelse(BikeAge_Gr %in% c("10-Jun","15-Nov"),NA,BikeAge_Gr) )
	#
	testFixReturn <- testFix( expression(length(rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value %in% c("10-Jun","15-Nov"),"freq"])), FALSE, as.integer(0) )
	if (testFixReturn[[1]]) {
		print("correct: zero rows remaining")
	} else {
		stop( paste(testFixReturn[[2]],testFixReturn[[3]]) )
	}
}
```

## Step 12. Noted by observation:   
There may be some problems with Crash_Loc and Crash_Type.
 
```{r step12, error=FALSE}
temp1 <- rawData_catcount[rawData_catcount$variable=="Crash_Loc" & rawData_catcount$value=="Location","freq"]
if ( temp1 > 0 ) {
	print( paste("There are", temp1, "instances of Crash_Loc = 'Location'. Set these to NA") )
	rawDataFixedMiss <- rawDataFixedMiss %>% dplyr::mutate( Crash_Loc=ifelse(Crash_Loc=='Location',NA,Crash_Loc) )
	testFixReturn <- testFix( expression(nrow(dplyr::filter(rawDataFixedMiss,Crash_Loc=='Location'))),
						TRUE, as.integer(0) )
	if (testFixReturn[[1]]) {
		print("correct: zero Crash_Loc=='Location' rows remaining in rawDataFixedMiss")
	} else {
		stop( paste("rawDataFixedMiss$Crash_Loc=='Location':",testFixReturn[[2]],testFixReturn[[3]]) )
	}
}
temp2 <- rawData_catcount[rawData_catcount$variable=="Workzone_I" & rawData_catcount$value=="No0","freq"]
if ( length(temp2) > 0 ) {
	print( paste("There are", temp2, "instances of Workzone_I = 'No0'. Set these to NA") )
	rawDataFixedMiss <- rawDataFixedMiss %>% dplyr::mutate( Workzone_I=ifelse(Workzone_I=='No0',NA,Workzone_I) )
	testFixReturn <- testFix( expression(nrow(dplyr::filter(rawDataFixedMiss,Workzone_I=='No0'))),
						TRUE, as.integer(0) )
	if (testFixReturn[[1]]) {
		print("correct: zero Workzone_I=='No0' rows remaining in rawDataFixedMiss")
	} else {
		stop( paste("rawDataFixedMiss$Workzone_I=='No0':",testFixReturn[[2]],testFixReturn[[3]]) )
	}
}
```

## Step 13. Recompute ancillary data frames and check on changes  
```{r step13, error=FALSE}
 rawDataFixedMiss_cat <- dplyr::select( rawDataFixedMiss, one_of(catcol_rawData) )
 rawDataFixedMiss_catmelt <- reshape2::melt(rawDataFixedMiss_cat,id="OBJECTID")
 rawDataFixedMiss_catcount <- plyr::count( rawDataFixedMiss_catmelt, vars=c("variable","value") )
 nrow(rawDataFixedMiss_catmelt)
 nrow(rawDataFixedMiss_catcount)
 # check to see that changes in previous step propagated correctly
 temp3 <- nrow( dplyr::filter(rawDataFixedMiss_catcount, variable=="BikeAge_Gr" & value %in% c("10-Jun","15-Nov")) )
 temp4 <- nrow( dplyr::filter(rawDataFixedMiss_catcount, variable=="Crash_Loc" &value=="Location") )
 temp5 <- nrow( dplyr::filter(rawDataFixedMiss_catcount, variable=="Workzone_I" & value=="No0") )
if (temp3+temp4+temp5 > 0) {
	stop( paste(temp1,temp2,temp3,temp4,temp5, "PROBLEM with setting BikeAge_Gr or Crash_Loc or Workzone_I entries to NA.") )
	} else {
	print("checks on ancillary data frames successful")
}
```

## Step 14. Write the csv file and save the session  
The files will be in your working directory - use getwd() to see what that is
```{r step14, error=FALSE}
write.table(rawDataFixedMiss, "rawDataFixedMiss_MD.csv", sep="\t")
save.image("bicycleSafetyData_MD.RData")
```
## FINIS
