---
title: "Durham NC Bicycle Safety data CodeForAmerica project"
author: "David Hopp "
date: "2016-05-09 "
output: html_document
---  

R markdown script to perform all operations  
leading to the final output csv file rawDataFixedMiss.csv  
 
## Step 0. Make sure required packages loaded
```{r step00, results="hide"}
require(plyr)
require(dplyr)
require(stringr)
require(reshape2)
```

## Step 1. Read the csv file acquired from the github project  
set "download" TRUE if file to be read from remote host  
set "download" FALSE if skipping remote read and using local csv file  
```{r step01}
# borrowed from Brian:
#
download <- FALSE
## entire path including file name
pathToCrashData <- "C:/Users/David Hopp/CodeForAmerica/bicycleData/scripts/RawDataFromRemote.csv" 
#
if (download) {
  strDownload <- "CrashData.csv.gz"  
  bikeSafetyURL <- "https://github.com/BikeSafety/BikeSafety/blob/master/data/North_Carolina_Bicycle_Crash_Data.csv.gz?raw=true"
  download.file( bikeSafetyURL, strDownload, mode="wb" )
  rawData <- read.csv( gzfile(strDownload), stringsAsFactors = FALSE )
  unlink(strDownload)
  rawData %>% write.csv( "RawDataFromRemote.csv", row.names = FALSE )
} else {
	if ( !exists("pathToCrashData") | is.na(pathToCrashData) | nchar(str_trim(pathToCrashData))==0 ) {
		stop( "Local file name not provided." )
	}
	if ( !file.exists(pathToCrashData) ) {
		stop( paste("Local file does not exist:", pathToCrashData) )
	}
	# at last
	rawData <- read.csv( pathToCrashData, sep=",", stringsAsFactors=FALSE )
}
#
nrow(rawData)
str(rawData)
```

## Step 2. Look at some data values
```{r step02}
 plyr::count(rawData, vars="BikeAge_Gr")
# notice the erroneous values at lines 3 and 4
```

## Step 3. Look at categorical data and see if any "odd" entries
```{r step03}
# As a result of observation the rawData categorical variables (columns) - keeping row number OBJECTID - are:
catcol_rawData <- c("OBJECTID", "AmbulanceR","BikeAge_Gr","Bike_Alc_D","Bike_Dir","Bike_Injur","Bike_Pos",
"Bike_Race","Bike_Sex","City","County","CrashAlcoh","CrashDay","Crash_Grp","Crash_Hour",
"Crash_Loc","Crash_Mont","Crash_Type","Crash_Ty_1","Crash_Year","Crsh_Sevri","Developmen","DrvrAge_Gr",
"Drvr_Age","Drvr_Alc_D","Drvr_EstSp","Drvr_Injur","Drvr_Race","Drvr_Sex","Drvr_VehTy","ExcsSpdInd","Hit_Run",
"Light_Cond","Locality","Num_Lanes","Num_Units","Rd_Charact","Rd_Class","Rd_Conditi",
"Rd_Config","Rd_Defects","Rd_Feature","Rd_Surface","Region","Rural_Urba","Speed_Limi","Traff_Cntr","Weather",
"Workzone_I" )
#
# work with only the categorical columns
rawData_cat <- dplyr::select( rawData, one_of(catcol_rawData) )
#
str(rawData_cat)
 # compute the values and counts of the categorical columns
 # these will require visual review to identify any odd entries
 #
 rawData_catmelt <- reshape2::melt(rawData_cat,id="OBJECTID")
 nrow(rawData_catmelt)
 rawData_catcount <- plyr::count( rawData_catmelt, vars=c("variable","value") )
 nrow(rawData_catcount)
```
 
## Step 4. Observations  

  Handle BLANKS leading and trailing  (will use str_trim to remove them)  
 
 There may be some entries that have trailing blanks  
 one way to find those is here:
```{r step04}
 rawData_catmelt_trblank <- dplyr::filter(rawData_catmelt, str_detect(value," +$"))
 nrow(rawData_catmelt_trblank)
 # which var has trailing blanks?
 unique(rawData_catmelt_trblank$variable)
 #
 # some may start with a blank
 rawData_catmelt_strblank <- dplyr::filter(rawData_catmelt, str_detect(value,"^ +"))
 nrow(rawData_catmelt_strblank)
 #
 reshape2::dcast(rawData_catmelt_strblank,value~variable)
```

# Here are some significant observations:  
There are 4 definitely *undependable* records in rawData: 
				2902, 3188, 4108, 4620  
There are other records that look _unreliable_:  
				371, 846, 942, 1247, 1271, 1777, 2101, 2130, 2789  
Note that BikeAge_Gr has
 `r rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value=="10-Jun","freq"]` instances of "10-Jun" and 
 `r rawData_catcount[rawData_catcount$variable=="BikeAge_Gr" & rawData_catcount$value=="15-Nov","freq"]` instances of "15-Nov"  
   

## Step 5. There are records with *empty* columns or pseudo-codes for missing values:  
(same for trimmed value)
```{r step05}
rawData_catcount[rawData_catcount$value=="",]
# (same for trimmed value)
rawData_catcount[rawData_catcount$value=="/Missing",]
# (same for trimmed value)
rawData_catcount[rawData_catcount$value=="Missing",]
```
Note that Crash_Loc has `r rawData_catcount[rawData_catcount$variable=="Crash_Loc" & stringr::str_trim(rawData_catcount$value)=="Location","freq"]` entries with "Location"  

## Step 6. "Fix" rawData  
result will be df **rawDataFixedMiss**  

 SPECIAL TREATMENT OF COLUMNS Longitude and Location  
 zeroes appear in both, as 0 and 0.00000 for Longitude  
 and (0, 0) for Location (watch out for the space)  
 these are best replaced with NA but that depends on   
 how Location is handled in the subsequent processing programs.  
 For now, replace with NA  
  
```{r step06}
rawDataFixed <- rawData
nrow(rawDataFixed)
# remove suspicious rows 
rawDataFixed <- rawDataFixed[ -c(2902, 3188, 4108, 4620,371, 846, 942, 1247, 1271, 1777, 2101, 2130, 2789), ,
       drop=FALSE]
nrow(rawDataFixed)
# TRIM TEXT FIELDS
mycols <- c("Bike_Injur","Crash_Loc","Crash_Type","Crsh_Sevri","Drvr_Injur","Drvr_VehTy")
rawDataFixed[mycols] <- lapply( rawDataFixed[mycols], function(x) stringr::str_trim(x) )
nrow(rawDataFixed)
```

## Step 7. Go through prep steps for categorical variables  
using rawDataFixed as the basis  
```{r step07}
rawDataFixed_cat <- dplyr::select( rawDataFixed, one_of(catcol_rawData) )
#
 rawDataFixed_catmelt <- reshape2::melt( rawDataFixed_cat,id="OBJECTID" )
 rawDataFixed_catcount <- plyr::count( rawDataFixed_catmelt, vars=c("variable","value") )
 #
 nrow(rawDataFixed_catmelt)
 nrow(rawDataFixed_catcount)
 rawDataFixed_catmelt_trblank <- dplyr::filter( rawDataFixed_catmelt, str_detect(value," +$") )
 #
 # the result should be zero rows:
 #
 nrow(rawDataFixed_catmelt_trblank)
```

## Step 8. Fix missing values  
replace with NA  
```{r step08}
 rawDataFixedMiss <- rawDataFixed
 rawDataFixedMiss[rawDataFixedMiss=="" | rawDataFixedMiss=="." | rawDataFixedMiss=="Missing" | rawDataFixedMiss=="/Missing"] <- NA
 nrow(rawDataFixedMiss)
```

## Step 9. Fix Longitude and Location - replace zeros with NA   
```{r step09}
 sum(rawDataFixedMiss$Location=="(0, 0)")

 rawDataFixedMiss$Location[rawDataFixedMiss$Location=="(0, 0)"] <- NA
 sum(is.na(rawDataFixedMiss$Location))
 sum(rawDataFixedMiss$Longitude==0)
 rawDataFixedMiss$Longitude[rawDataFixedMiss$Longitude==0] <- NA
 sum(is.na(rawDataFixedMiss$Longitude))
```

## Step 10. Fix BikeAge_Gr when it contains "10-Jun" or "15-Nov"
```{r step10}
 rawDataFixedMiss$BikeAge_Gr[rawDataFixedMiss$BikeAge_Gr %in% c("10-Jun","15-Nov")] <- NA
```

## Step 11. Noted in passing:   
There are `r rawData_catcount[rawData_catcount$variable=="Workzone_I" & rawData_catcount$value=="No0","freq"]` Workzone_I entries with "No0" instead of "No"   
`r rawData_catcount[rawData_catcount$variable=="Crash_Loc" & rawData_catcount$value=="Location","freq"]` Crash_Loc with "Location"
`r rawData_catcount[rawData_catcount$variable=="Crash_Type" & rawData_catcount$value=="Location","freq"]` Crash_Type with "Location" 
```{r step11}
 rawDataFixedMiss[rawDataFixedMiss=="Location"] <- NA
 rawDataFixedMiss[rawDataFixedMiss=="No0"] <- "No"
```

## Step 12. Look closely at the categorical variables
```{r step12}
 rawDataFixedMiss_cat <- dplyr::select( rawDataFixedMiss, one_of(catcol_rawData) )
 rawDataFixedMiss_catmelt <- reshape2::melt(rawDataFixedMiss_cat,id="OBJECTID")
 rawDataFixedMiss_catcount <- plyr::count( rawDataFixedMiss_catmelt, vars=c("variable","value") )
 nrow(rawDataFixedMiss_catmelt)
 nrow(rawDataFixedMiss_catcount)
```

## Step 13. Write the csv file and save the session
```{r step13}
# The files will be in your working directory - use getwd() to see what that is
#
write.table(rawDataFixedMiss, "rawDataFixedMiss_MD.csv", sep="\t")
save.image("bicycleSafetyData_MD.RData")
```
## FINIS
